---
title: "Working with Spatial Data: Tutorial"
author: "Joe Celebrezze"
date: "2025-05-26"
output: html_document
---

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# highly necessary packages
library(tidyverse)
library(terra) # to work with rasters (mostly)
library(sf) # to work with shapefiles (mostly)

# extra packages that we use today
library(here) # handy package to simplify file paths
library(tidyterra) # easily implement terra objects into tidyverse/ggplot
library(tmap) # nice visualizations for maps

# other packages that folks use (that we won't use today)
# library(stars) # an alternative package to work with spatial data
# library(raster) # an outdated package to work with raster data that is slower than terra (I still have this one in a lot of my scripts, but I am working towards phasing it out)
# library(mapview) # an alternative package to make interactive maps
# library(leaflet) # another package to make interactive maps
# library(gstat) # for spatial stats
# library(rgbif) # to download species occurrence data from gbif
# library(geodata) # to download geographic data including climate, elevation and species occurrence
```

# Downloading Data
```{r}
# First, downloading some shapefiles
willow_pts <- read_sf(here('data', 'willow_pts.shp')) # coordinates of sites
replanted_polygons <- read_sf(here('data', 'willow_FACTS_replanting.shp')) # polygons of areas where USFS planted trees post-fire

# Next, downloading some rasters
burn_severity <- rast(here('data', 'willow_2015_dnbr6.tif')) # MTBS burn severity classes for Willow fire
nbr_stack <- rast(here('data', 'willow_NBR.tif')) # raster stack of growing season (June 1-August 31) NBR from 1998-2024
```

# Coordinate Systems and Extents
```{r}
# To check what coordinate system our spatial data is stored in
st_crs(burn_severity)

# Holy cow that is a lot... To just check out a short-form of CRS, we can:
st_crs(burn_severity)$Name

# Are all spatial data in the same CRS?
st_crs(nbr_stack)$Name
st_crs(willow_pts)$Name
st_crs(replanted_polygons)$Name

# Reprojecting spatial data so they're all in the same CRS
# using terra (for rasters)
nbr_stack <- project(nbr_stack, crs(burn_severity))
st_crs(nbr_stack)$Name
# using sf (for shapefile)
replanted_polygons <- st_transform(replanted_polygons, st_crs(burn_severity))
st_crs(replanted_polygons)$Name
```

What are the extents of these spatial data?
```{r}
ext(nbr_stack)
ext(burn_severity)
# They're a bit different
```

# Simple Maps
(as simple as it gets)
```{r}
# select both and run or run chunk
plot(burn_severity) # MTBS burn severity classes
points(willow_pts) # adding our field sites

# NBR from 2014-2017 (fire in later 2015)
plot(nbr_stack[[17:20]])

# alternatively, to only visualize years 2014-2017 we could subset those years using terra::subset()
nbr_subset <- subset(nbr_stack, c(17:20))
plot(nbr_subset)
```

Looking at the maps of the burn severity and the nbr stack, we can see why their extents were slightly different, even though they're the same fire

# Extracting Data
The most common thing I do with spatial data in R is... change it so that it's not spatial data. Although it can be very nice to retain the spatial information, it can also make code take longer to run and it can be sometimes unnecessary. If we only want the information encoded in the rasters or shapefiles for our sites, then we can extract the data and work with that dataframe. 
```{r}
# from rasters
severity_vals <- extract(burn_severity, willow_pts) # burn severity
class(severity_vals) # stored as a dataframe
colnames(severity_vals)[2] <- 'burn_severity' # to rename column
nbr_vals <- extract(nbr_stack, willow_pts) # yearly NBR

# from shapefile
treatment_vals_all <- st_join(willow_pts, replanted_polygons) # option 1: for all of the points, add information from the polygon; if they don't intersect with polygon, add NAs; note, if polygons intersect, it will add rows for each polygon that the point is in
class(treatment_vals_all) # stored as a spatial dataframe (simple features)
treatment_vals_intersects <- st_join(willow_pts, replanted_polygons, left = F) # option 2: remove any rows where points do not intersect with polygons

# clean up treatment_vals_all
clean_treatment_vals <- treatment_vals_all %>% 
  select(id.x, geometry, ACTIVITY) %>% 
  rename(ID = id.x) %>%  # because there was an id column in replanted_polygons, we had id.x and id.y in resulting spatial dataframe
  mutate(count = ifelse(is.na(ACTIVITY), 0, 1)) # add a column where we have a 1 for each instance of replanting trees

clean_treatment_sums <- clean_treatment_vals %>% 
  group_by(ID) %>% 
  summarise(replanting = sum(count)) %>%  # so that new dataframe has a column for each site equal to the number of times the area was replanted
  st_drop_geometry() # removing spatial component

# combining dataframes
extracted_df <- severity_vals %>% 
  left_join(clean_treatment_sums, by = 'ID') %>% 
  left_join(nbr_vals, by = 'ID')
```

# Visualizations
## Time Series
```{r}
# first, get the data in long format where we have a column for year and a column for NBR
long_df <- extracted_df %>% 
  pivot_longer(cols = !c('ID', 'burn_severity', 'replanting'),
      names_to = 'year',
      values_to = 'nbr') %>%
  mutate(burn_severity = case_when(
    burn_severity == 1 ~ 'Unburned',
    burn_severity == 2 ~ 'Low Severity',
    burn_severity == 3 ~ 'Moderate Severity',
    burn_severity == 4 ~ 'High Severity')) %>%  # so that our burn severity values are a bit more informative
  mutate(year = as.numeric(year)) # so that it's not stored as a character vector

# for each burn severity class, take the median value of NBR for each year
medians_df <- long_df %>% 
  group_by(burn_severity, year) %>% 
  summarise(nbr = median(nbr))

# visualization
ggplot() +
  geom_line(aes(x = year, y = nbr), data = medians_df, linewidth = 2) +
  geom_point(aes(x = year, y = nbr), data = long_df, alpha = 0.3) +
  geom_vline(xintercept = 2015, linetype = 'dashed') +
  facet_wrap(~burn_severity) +
  labs(x = 'Year', y = 'Normalized Burn Ratio') +
  theme_classic()
```

## dNBR vs. MTBS Severity
```{r}
extracted_df %>% 
  mutate(burn_severity = case_when(
    burn_severity == 1 ~ 'Unburned',
    burn_severity == 2 ~ 'Low Severity',
    burn_severity == 3 ~ 'Moderate Severity',
    burn_severity == 4 ~ 'High Severity')) %>%  # so that our burn severity values are a bit more informative
  mutate(burn_severity = fct_relevel(burn_severity, c('Unburned', 'Low Severity', 'Moderate Severity', 'High Severity'))) %>%  # so that the x axis is in the order we want
  mutate(dnbr = `2015`-`2016`) %>%  # calculating dNBR
  ggplot(aes(x = burn_severity, y = dnbr)) +
    geom_boxplot(outlier.shape = NA, fill = NA, ) +
    geom_jitter(width = 0.2, height = 0) +
    labs(x = 'MTBS Burn Severity', y = 'dNBR') +
    theme_bw() +
    theme(panel.grid = element_blank())
```


Okay, that was a bit of a digression. Let's get back to playing with our spatial data.

# Raster Math
NIRv is an index that is more closely related to photosynthesis and gross primary productivity (GPP) and it is equal to NDVI*NIR. However, I use LandTrendr in Google Earth Engine to get my time series of remotely sensed data and they don't have an option to directly get time series for NIRv. They do have NDVI and NIR though! We can simply multiply them here.
```{r}
ndvi_stack <- rast(here('data', 'willow_NDVI.tif')) # raster stack of growing season (June 1-August 31) NDVI from 1998-2024
nir_stack <- rast(here('data', 'Willow_NIR.tif')) # raster stack of growing season (June 1-August 31) NIR from 1998-2024

nirv_stack <- ndvi_stack*nir_stack # calculating NIRv

# checking out NIRv from 2014-2017
plot(nirv_stack[[17:20]])
```

# Nicer Maps
## ggplot
```{r}
# another (more complicated) example of raster math
RdNBR <- (nbr_stack[[19]]-nbr_stack[[18]])/(sqrt(mean(nbr_stack[[1:17]]))) # change in NBR due to fire divided by the square root of the pre-fire mean

ggplot() +
  geom_spatraster(data = RdNBR) + # adding raster (easy with tidyterra)
  scale_x_continuous(guide = guide_axis(check.overlap = T)) + # declutter x axis
  labs(fill = 'RdNBR') + # legend title
  scale_fill_gradientn(colors = c('black', 'white', 'darkgreen')) + # change color palette
  theme_minimal() +
  theme(panel.grid = element_blank()) # clean up plot
```

What if we want to add polygons and rasters?

geom_spatvector for vectors
geom_spatraster for rasters

```{r}
# downloading boundary
willow_boundary <- read_sf(here('data', 'willow_boundary.shp')) # coordinates of sites

ggplot() +
  geom_spatraster(data = RdNBR) + # adding raster
  geom_spatvector(data = replanted_polygons, color = 'black',
                  fill = 'cyan3', alpha = 0.5) + # adding polygons for planted areas
  geom_spatvector(data = willow_boundary, color = 'black',
                  fill = NA, linewidth = 1) + # adding polygon for boundary
  geom_spatvector(data = willow_pts, aes(color = 'Field Site')) + # adding polygon for points
  scale_color_manual(name = "Points",
    values = c("Field Site" = "black")) + # to add sites to legend
  scale_x_continuous(guide = guide_axis(check.overlap = T)) + # declutter x axis
  labs(fill = 'dNIRv') + # legend title
  scale_fill_gradientn(colors = c('black', 'white', 'darkgreen')) + # change color palette
  theme_minimal() +
  theme(panel.grid = element_blank()) # clean up plot
```

## tmap
In my experience, interactive maps like tmap or leaflet can be relatively clunky on R and it can be challenging to add everything you want to add. If I want something more interactive, I often use QGIS or ArcGIS.
```{r}
# To have an interactive map
tmap_mode("view")

# Adding elements to the map
# burn severity raster
tm_shape(burn_severity) +
  tm_raster(style = "cat",  # treat as categorical
    palette = c(
      "0" = "#FFFFFF00",  # null values (transparent)
      "1" = "#006400",   # Unburned 
      "2" = "#7fffd4",   # Low severity
      "3" = "#ffff00",      # Moderate severity
      "4" = "#ff0000",   # High severity
      "5" = "#FFFFFF00"),     # Increased grenness (transparent)
    title = "Burn Severity") + # note the plus sign to continue adding elements to tmap
  
# Replanted polygons
tm_shape(replanted_polygons) +
  tm_polygons(
    col = "cyan3",
    border.col = "black",
    alpha = 0.5,
    legend.show = TRUE,
    id = "Replanted Area") +

# Field sites
tm_shape(willow_pts) +
  tm_symbols(
    col = "black",
    shape = 21,
    size = 0.1,
    legend.col.show = TRUE,
    id = "Field Site")
```

```{r}
# To have an static map
tmap_mode("plot")

# Adding elements to the map
# burn severity raster
tm_shape(burn_severity) +
  tm_raster(style = "cat",  # treat as categorical
    palette = c(
      "0" = "#FFFFFF00",  # null values (transparent)
      "1" = "#006400",   # Unburned 
      "2" = "#7fffd4",   # Low severity
      "3" = "#ffff00",      # Moderate severity
      "4" = "#ff0000",   # High severity
      "5" = "#FFFFFF00"),     # Increased grenness (transparent)
    title = "Burn Severity") + # note the plus sign to continue adding elements to tmap
  
# Replanted polygons
tm_shape(replanted_polygons) +
  tm_polygons(
    col = "cyan3",
    border.col = "black",
    alpha = 0.5,
    legend.show = TRUE,
    id = "Replanted Area") +

# Field sites
tm_shape(willow_pts) +
  tm_symbols(
    col = "black",
    shape = 21,
    size = 0.1,
    legend.col.show = TRUE,
    id = "Field Site")
```

# Other Useful Stuff
Making stacks and renaming rasters
```{r}
# to make a raster stack using terra, we can simply use c()
dNBR <- nbr_stack[[19]]-nbr_stack[[18]] # to calculate dNBR
homemade_stack <- c(RdNBR, dNBR)
names(homemade_stack) <- c('RdNBR', 'dNBR') # to rename the rasters
plot(homemade_stack)
```

Classifying cell values
```{r}
# if we want to use an existing raster to define a new raster (similar to ifelse or case_when), we can use classify()
# first, set a matrix to use for the classification scheme; we will use dNBR values to classify burn severity
classification_matrix <- matrix(
  c(-100, 350, 1, # Unburned
    -270, -100, 2, # Low
    -450, -270, 3, # Moderate
    -1000, -450, 4), # High
    ncol = 3, byrow = T)

# Now, we can use that matrix to classify dNBR
homemade_burn_severity <- classify(dNBR, classification_matrix, include.lowest = T)

# compare to mtbs
plot(homemade_burn_severity)
plot(burn_severity)
```

## Sampling Points
```{r}
# uniform sampling
regular_pts <- spatSample(RdNBR, 54, method = 'regular') # only return values

regular_spatial_pts <- spatSample(RdNBR, 54, method = 'regular', as.raster = T) # return points

# let's check it out
plot(RdNBR)
points(regular_spatial_pts)
# note how ~20 points are outside of the plot because it uses the raster extent to set the bounds of spatSample

# random sampling
random_spatial_pts <- spatSample(burn_severity, 10, method = 'random', as.raster = F, xy = T)
# check it out
plot(burn_severity)
points(random_spatial_pts, pch = 21, bg = 'blue', col = 'black')
```

To be honest, it's not all-that-helpful to sample points within the extent of the raster. In this case, we'd prefer to sample points within the extent of the burn boundary. So, let's use sf instead of terra
```{r}
sf_random_pts <- st_sample(willow_boundary, 10, type = 'random')

plot(burn_severity)
points(sf_random_pts, pch = 21, bg = 'blue', col = 'black')
# much better!
```
